---
title: "GIS_workshop1_Nowosad_Lovelace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GIS_workshop1_Nowosad_Lovelace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Slides: https://nowosad.github.io/SIGR2021/workshop1/workshop1_rl.html#1

# Example datasets
## Why use example Datasets

- learning and documenting the use of the package
- use a small dataset rather than a large dataset => quicker 
- unit testing

size of a sample dataset:
- small enough to be fast
- but with edge cases

You can have multiple datasets (small/large, differents areas)

Robin recommends to create an example dataset to save time.
Documenting it is recommended but not mandatory.


ActDev example: know where to stop R: R is the backend, the webapplication to visualise the data is in JS.

```{r R-datasets}
library(help ="datasets") # full list of datasets
```

Listed datasets are available directly from R.

### Plot 2 datasets
```{r plot-trees}
plot(trees)
```

```{r plot-Lake-Huron}
plot(LakeHuron)
```

### Get information on data in a package

```{r}
help(package = "spData")
```

```{r airpassenger-help}
?AirPassengers
```

```{r plot-nz}
library(spData)
plot(nz) # if list error, maybe because sf is not loaded yet and plot don't know how to plot sf objects

plot(nz$geom)
```


Make a copy to work on is good practice.

```{r make-a-copy}
nz2 = nz
plot(nz)
```

```{r}
library(nzelect)

nz_lonlat = sf::st_transform(nz, 4326)
names(voting_places)

# create points from coordinates
voting_places_sf = sf::st_as_sf(voting_places, coords = c("longitude", "latitude"))

plot(sf::st_geometry(nz_lonlat))
plot(voting_places_sf, add = TRUE)
```

### Spatial subsetting

Filtering voting places that are inside NZ island
```{r}
vsub = voting_places_sf[nz,] # Error message because mismatching CRS issues
```

```{r}
sf::st_crs(voting_places_sf) # NA
sf::st_crs(nz) # EPSG:2193
```

Voting_places_sf does not have a crs

Fixing issues

```{r}
voting_places_sf <- sf::st_set_crs(voting_places_sf, st_crs(4326))
sf::st_crs(voting_places_sf)
nz_lonlat <- sf::st_transform(crs = st_crs(4326))
vsub = voting_places_sf[nz_lonlat,]
```


```{r}
plot(sf::st_geometry(nz_lonlat))
plot(vsub, add = TRUE)
```

# Using OSM data

Quite informal but very very rich
Interesting when you can't access official data


Simple requests: {osmdata}

```{r}
library(osmdata)
```

```{r}
schools_nz_osm <- opq(bbox = sf::st_bbox(nz_lonlat)) %>%
  add_osm_feature(key = "amenity", value = "school")
```

## Working with messy data

`skimr:skim` : resume data

- selecting variables of interest : saving space and time



### source outside code
```{r}
source("https://raw.githubusercontent.com/Nowosad/SIGR2021/master/workshop1/osm_combine.R")

# osm_combine() is available in the environnment
```

## Big OSM data

For large dataset (> 100 MB)
```{r}
library(osmextract)
```

Needs a place name, and parameters (date type (e.g. "amenity") and an SQL query for the OSM server)

# Tidyverse

```{r}
library(spData)
library(sf)
library(tidyverse)

nz %>%
  filter(Island == "South")
```

## Geometry operations

- Simplification : `rmapshaper::ms_simplify`
- `st_is_valid()` : check if geometries are valid
- `st_make_valid()` : fixes geometries

## download online data
- `download.file`
- `readr` can read some ressources
## 
