---
title: "Image processing and all things raster"
author: "Jakub Nowosad"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GIS_workshop3_image_processing_Nowosad}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Slides : https://nowosad.github.io/SIGR2021/workshop2/workshop2.html#1

```{r data-input}
srtm_path = system.file("raster/srtm.tif", package = "spDataLarge")
srtm_path
```

`system.file` : nice way to get access to example data contained in packages by getting the path.

R packages are better for storing files than R objects. FIles are mor generic and broader.


Load {terra} and the dataset
```{r}
library(terra)
srtm = rast(srtm_path)
srtm
```

`rast()` just connect to the file but not loading into memory.
It also provide useful information about the dataset.



Looks to alternatives of {terra}:
- {raster} : predecessor of {terra}, no new developpement
- {stars} : compatible with {sf}, quick, multidimensional, support different types of raster (regular (90% of use cases), rotated, sheared, rectilinear, curvilinear), connect to geospatial servers (works in progress)

# Maps
## Mapping tools
Maps were detailed earlier, this workshop will use `plot()` and `tmap()`

Generally, `plot()` functions simplifies datasets larger than the image dimensions.
It is faster to do that.

With `tmap()` if there is more than 1 million cells, the data will bo down sampled by `tmap()`

# Map algebra

Four groups
- **local** : computation one cell
- **focal** : cell value and its neigbourg (*moving window*)
- **zonal operations** summarize raster values for somes zones
- **global** summarize the whole raster (max, min ...)

## Local operations

- raster calculator
 - `srtm2 = srtm * 1000`
 - `srtm3 = srtm - globale(srtm, min)[[1]]` => 0 -> to max value
 - `srtm3 = srtm - globale(srtm, median)[[1]]` => 0 will be the median value, negative value are under the median value
- replacing values
```{r}
srtm_new = srtm
srtm_new[srtm_new < 1500] = NA # values under 1500 will be NA
```

- reclassification
  - Reclassification table (2 or 3 columns) : from this value to this value replace by this value
  - `classify` takes reclassification table as argument
- operations on many layers (calculating spectral indices, NDVI and such)
  - landsat dataset is a simplified dataset from landsat
  
```{r}
landsat_path = system.file("raster/landsat.tif", package = "spDataLarge")
landsat = rast(landsat_path)
landsat
```

  - Write a function to compute the ndvi.
  - use the `lapp()` function
```{r}
ndvi_fun = function(nir, red){
  (nir - red) / (nir + red)
}
ndvi = lapp(landsat[[c(4, 3)]], # subset 2 layers (4 and 3) from the dataset
            fun = ndvi_fun) # custom ndvi function
```




## Focal operations

Works on cell neigbourhoood, we need to provide the neighbourghood (*window*)

The window is provided by a matrix (mandatory type).
The function will be applied on the center cell.
The matrix can be weighted to have special effect.
The matrix needs to be uneven 3x3, 5x5, 7x7. With 4x4 matrix, the function will not be able to determined the center.

**Sobel operator** : edge detector

The operation starts from the top left to the bottom right, but the order does not change the output.


You can use focal operation on categorical raster : "what's the most dominant value."

## Zonal operations

also know as zonal statistics.
Results are **summary table** not a raster.

You can provide a function, some are optimised : "mean", "max", "min", "media" **note the `"`**.

## Global operation

Got a result for the whole raster.
`global()`

## Exercices


- Find the file path of the nz_elev.tif data using the following code: system.file("raster/nz_elev.tif", package = "spDataLarge"). Read this file into R. What is the resolution of the data?

```{r}
system.file("raster/nz_elev.tif", package = "spDataLarge")

nz_elev <- terra::rast(system.file("raster/nz_elev.tif", package = "spDataLarge"))
nz_elev
```

The resolution is `resolution  : 1000, 1000  (x, y)`

- Create and customize a map based on the nz_elev.tif dataset. Save the result to a .png file.

```{r}
library(tmap)

nz_map <- tm_shape(nz_elev) +
  #tm_graticules() +
  tm_raster(style = "cont", 
            title = "elevation (m a.s.l)",
            palette = "-Spectral") +
  tm_scale_bar(breaks = c(0, 2, 4),
               text.size = 1) +
  tm_credits("Nicolas Roelandt, 2021") +
  tm_layout(inner.margins = 0,
    main.title = "New Zealand")

tmap_save(nz_map, "vignettes/images/new_zealand_elevation.png")
```



- Replace values below 0 to 1 (depressions), values between 0 and 300 to 2 (plains), values between 300 and 600 to 3 (hills), and values between 600 and the maximum value to 4 (mountains) in the nz_elev object. Create a new object nz_elev_class.

```{r}
rcl = matrix(c(-Inf, 0, 1, 
               0, 300, 2, 
               300, 600, 3, 
               600, 4140.333, 4),
             ncol = 3, byrow = TRUE)
rcl
nz_elev_class <- classify(nz_elev, rcl = rcl)
nz_elev_class 
```

- Look at the focal function documentation. Apply the Laplacian and the Sobel filters on nz_elev. Compare the results visually.

- Calculate an average elevation value in nz_elev for each category in nz_elev_class.

- Write a function to calculate the soil-adjusted vegetation index (SAVI; https://en.wikipedia.org/wiki/Soil-adjusted_vegetation_index ). Calculate SAVI for the Landsat 8 images for the area of Zion National Park.


# Transformations
## Resampling
## Reprojecting rasters

- different from vector
- needs to resample
  - create an empty raster template
  - to fillthe empty value, use `resample()`
  - parameters : base image, empty raster, function to explain how to resample (method : "bilinear" - good for continues values, "nearest neighbourg" - only one for categorical data)
  
```{r}
wgs84 = "EPSG:4326"

nlcd_wgs84 = project(nlcd, wgs84, method = "near")
```
  
To create your own projection : [projectionwizard.org](https://projectionwizard.org/)

# Raster-vector interactions

## Raster cropping and masking
Cropping : remove data outside the BBox
Crop is fast

The mask function replace all values outside the shape with NA.
Mask is always used with a cropped image to delete NA data outside the shape

## Raster extraction 
### With points
- gps points or anything.

For example, I want elevation of each point using the elevation raster.

`terra::extract()`

The extract is a dataframe with ID and value.

### With polygons

A polygons covers a lotof pixels. 
Extract each pixel values and give the ID of the polygons.

Useful to see the distribution, do some some statistics.

## Rasterization
Transform vector data to raster data

### Presence / Absence

- `rasterize` if there is a point in the cell, I get a value of one (TRUE/FALSE)

- `fun = length` : count of point in the cell
- `field = "vals", fun = sum` : sum "vals" values of all points inside the cell

## Vectorization

From raster to vector.

Often to create polygons on a categorical raster data

- create borders
- contours lines (elevation lines, precipitations lines, isolines) : `as.contour()` : you can provide a number of line to create (`nlines`) or elevation limits `z` , 

# Conversions
From spatial type/class to another

# Output

Save raster file

`writeRaster()` : depending on the file extension we can have different format
With {terra}, geoTIFF are compressed by default.

# Raster analysis

## Global or local autocorrelation

`autocor()` :
- by default, global autocorrelation
- default is 'Moran' but you can use 'Geary' autocorrelation
For elevation, high autocorrelation is expected

For local correlation
- provide a window 
  - matrix neighborhood
  - can be `queen` (8-n) or `rook` (4-n) neighborhood
- `global = FALSE`
- the output is a raster: find place with high or low autocorrelation

## Predictions

{terra} objects can be used on glm, randomForest or prcomp functions.

## Interpolations
`interpolate()`

We have points but we want areas.

Create a model from the points (splines, kriging, IDW)

Geostatistiques with {gstat}

`gstat::variogram()` distance between values and how values are dissimilear to each other.

## Segmentations

{supercells} : super-pixels : automatically find similar areas in visible imagerie

Return an {sf} dataframe with centroids coordinates and mean pixel value of each band of supercell.
