---
title: "GIS and mapping" 
author: 
  - Jakub Nowosad
  - Robin Lovelace
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GIS and mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Slides: https://nowosad.github.io/SIGR2021/workshop1/workshop1_jn.html#1

```{r}
library(sf)
library(terra)
library(tmap)
library(spData)
library(spDataLarge)
```

# Mapping

```{r}
nz_elev = terra::rast(system.file("raster/nz_elev.tif", package = "spDataLarge"))
plot(nz_elev)
```

Tmap functions starts with `tm_` prefix

`tm_borders` : outline empty polygons
`tm_legends` : create manual legends
`tm_layout` : overall look of the map (margin, frame, title, background color)

# Shapes and layers

## raster data

- `tm_raster` : elevation, pluviometry, etc.
- `tm_rgb` : satellite imagery with 3 layers dataset (visible, false colors)

```{r}
tm_shape(nz_elev) + # data what we want to plot
  tm_graticules() + # add graticules, lines in degrees (not working ?)
  #tm_grid() + # grid for projected data
  tm_raster( # we want to plot the shape as a raster
    title = "elevation as asl", # change legend title
    style = "cont", # discretization (échelle continue, sans classes)
    palette = "-Spectral" # palette spectal inversée
    #palette = c("green","blue", "yellow", "red") # palette custom
            ) +
  tm_shape(nz) + # New Zealand vector data
  tm_borders(col = "black", # black borders with empty polygons see tm_fill() and tm_polygons()
             lwd = 1 ) +
  tm_shape(nz_height) +
  tm_symbols(size = 0.2, col = 'red', shape = 17) + # chnage point size and colors
  tm_add_legend(type = "symbol", col = "red", shape = 17, title = "Highest peaks") +
  tm_scale_bar(breaks = c(0,100,200)) +
  tm_compass(position = c("right", "top"),
             type = "radar",
             size = 2) +
  tm_credits(text = "N. Roelandt, 2021") +
  tm_layout(main.title = "New Zealand topography",
            bg.color = "lightblue", # background color #add8e6
            inner.margins = c(0.01,0.01,0.01,0.2), # bottom, left, top, right
            legend.outside = TRUE,
            frame = FALSE)

  
```

```{r}
library(sf)

nz_4326 = st_transform(nz, 4326)

tm_shape(nz_4326) +
  tm_polygons(text.size = 1, breaks = c(0,20))
```


## Explore color palettes in {tmap}

```{r, eval=FALSE}
tmaptools::palette_explorer()# needs shiny installed
```

```{r}
tm_shape(nz) +
  tm_polygons(col = "Median_income",
              title = "Median income (USD)",
              style = "jenks",
              n = 20,
              legend.hist = TRUE) +
  tm_layout(inner.margins = c(0,0.02,0,0.5)) # ajust margins
```

```{r}
tm_shape(nz) +
  tm_polygons(col = "Median_income",
              title = "Median income (USD)",
              style = "jenks",
              n = 20,
              legend.hist = TRUE,
              legend.format = list(digits =1 )) + # mise en forme de la legend
  tm_layout(legend.outside = TRUE) # move legend outside the map
```


# Modes

Static map / interactive map

```{r}
tmap_mode("view") +
tm_shape(nz_elev) + # data what we want to plot
  tm_graticules() + # add graticules, lines in degrees (not working ?)
  #tm_grid() + # grid for projected data
  tm_raster( # we want to plot the shape as a raster
    title = "elevation as asl", # change legend title
    style = "cont", # discretization (échelle continue, sans classes)
    palette = "-Spectral" # palette spectal inversée
    #palette = c("green","blue", "yellow", "red") # palette custom
            ) +
  tm_shape(nz) + # New Zealand vector data
  tm_borders(col = "black", # black borders with empty polygons see tm_fill() and tm_polygons()
             lwd = 1 ) +
  tm_shape(nz_height) +
  tm_symbols(size = 0.2, col = 'red', shape = 17) + # chnage point size and colors
  tm_add_legend(type = "symbol", col = "red", shape = 17, title = "Highest peaks") +
  tm_scale_bar(breaks = c(0,100,200)) +
  tm_compass(position = c("right", "top"),
             type = "radar",
             size = 2) +
  tm_credits(text = "N. Roelandt, 2021") +
  tm_layout(main.title = "New Zealand topography",
            bg.color = "lightblue", # background color #add8e6
            inner.margins = c(0.01,0.01,0.01,0.2), # bottom, left, top, right
            legend.outside = TRUE,
            frame = FALSE)

```

# Saving

```{r}
my_map = tm_shape(nz_elev) + # data what we want to plot
  tm_graticules() + # add graticules, lines in degrees (not working ?)
  #tm_grid() + # grid for projected data
  tm_raster( # we want to plot the shape as a raster
    title = "elevation as asl", # change legend title
    style = "cont", # discretization (échelle continue, sans classes)
    palette = "-Spectral" # palette spectal inversée
    #palette = c("green","blue", "yellow", "red") # palette custom
            ) +
  tm_shape(nz) + # New Zealand vector data
  tm_borders(col = "black", # black borders with empty polygons see tm_fill() and tm_polygons()
             lwd = 1 ) +
  tm_shape(nz_height) +
  tm_symbols(size = 0.2, col = 'red', shape = 17) + # change point size and colors
  tm_add_legend(type = "symbol", col = "red", shape = 17, title = "Highest peaks") +
  tm_scale_bar(breaks = c(0,100,200)) +
  tm_compass(position = c("right", "top"),
             type = "radar",
             size = 2) +
  tm_credits(text = "N. Roelandt, 2021") +
  tm_layout(main.title = "New Zealand topography",
            bg.color = "lightblue", # background color #add8e6
            inner.margins = c(0.01,0.01,0.01,0.2), # bottom, left, top, right
            legend.outside = TRUE,
            frame = FALSE)

my_map

tmap::tmap_mode("view")
my_map
tmap::tmap_mode("plot")
my_map
```

```{r, eval=FALSE}
tmap_save(my_map, filename = "vignettes/images/nz.png", width = 300, height = 800, dpi = 300)
tmap_save(my_map, filename = "vignettes/images/nz.svg")
tmap_save(my_map, filename = "vignettes/images/nz.html")
```

* `tm_` : functions to make maps
* `tmap_` : interact with the whole map

# Exercices
```{r}
library(sf);library(tmap);library(spData);library(dplyr)
us_states = left_join(us_states, us_states_df, by = c("NAME" = "state"))
us_states2163 = st_transform(us_states, "EPSG:2163")
```



- Run the code above. Create a simple map of the reprojected data for contiguous United States (object us_states2163) with the graticule lines.

```{r}
us_map = tm_shape(us_states2163) + # data what we want to plot
  tm_grid() +
  tm_borders(col = "black", lwd =1)
us_map
```

- Improve the map of the reprojected data for contiguous United States by coloring states based on their median income in 2015 (median_income_15). Improve the legend title. Try different color palettes. Change the width of the borders. Put the legend on the right side of the map.

```{r}
us_map = us_map +
  tm_fill(col = "median_income_15",
          title = "2015 median income (in US$)",
          palette = )
  
us_map
```

- Try different color scale styles. Which one is the best in your opinion?
- Improve the map of the reprojected data for contiguous United States (object us_states2163) further by adding a map title, scale bar, north arrow, and text annotation. Try to improve locations of the map elements.
- Change the tmap mode to "view" and recreate maps from Exercise 4. What are the pros and cons of using static or interactive maps? What are the map's elements that can exist in static maps, but are not rendered in the interactive maps?
- Save the map from Exercise 4 as a .png, .svg, and .html file. When can different file formats be useful?

